evidence of adaption-independent time-warped spike timing noise correlations in cortical networks

### 2020-07-30 manuscript.tex
V2.0 - Introduction

\title{evidence of stationary and non-stationary clustered time-warped spike timing in cortical networks}

@@

\section{Introduction}
The neural coding problem asks how information is encoded through patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}, with a dichotomy between temporal \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar} and rate codes \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}. 
% Temporal codes provide the possibility of a fast and efficient code. 
Whilst precise temporal patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of neural variability remains unexplained. Such variability leads to the common assumption of average firing rate codes, under the premise that variability cancels out across a population \cite{abbott1999effect, moreno2014information, stringer2019high}. Continuous firing rates (averaged over time, trial or population \cite{gerstner1997neural}) have capacity beyond the single trial discrete spike count from which they are abstracted, leaving a full understanding of neural coding and variability undetermined. It is important to better characterise neural codes in order to understand a rich variety of dynamic, structural, homeostatic and cognitive processes, as well as properties of action potential patterns directly involved in behaviour.
% Characterising responses may also uncover response properties involved in behaviour.

Responses following sensory stimulation can be analysed to explore principles of neural coding. Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus. A multi-neuron response can be expressed as a point in multi-dimensional space, where each dimension represents a spike train metric. Responses from multiple presentations of a stimulus form a sample stimulus-response distribution in this space (Fig. 1a). 
% which approximates an underlying stimulus-response distribution.

@@

Information theoretic approaches aim to understand 
If two stimuli have a behaviourally relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus. This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig. 1a). Characterising the distribution from which a stimulus response is `drawn' from on a single trial is complicated by correlative structure, neural adaptation and heterogeneity of sample response distribution structure.

Correlative structure between the responses of different neurons in single stimulus response distributions indicates the presence of noise shared between the responses of neurons. These are referred to as `noise correlations' and are the focus of this manuscript, rather than `signal correlations' which refer to tuning similarity between neurons over multiple stimuli.

Information theoretic approaches classically consider neural encoding with respect to fixed-state encoding and decoding. These approaches often do not consider that correlative structure (i.e. covariance of multi-neuron responses over trials of a single stimulus) could be caused by a \textit{continuous} underlying state, which covaries neural responses from trial-to-trial.

Firstly, correlative structure between the responses of spatially distanced neurons (see \cite{reid2012functional, feldmeyer2013barrel}) in a single stimulus response distribution, opens the possibility of a \textit{continuous} underlying state, which covaries neural responses from trial-to-trial. Response samples on a single trial, when the continuous state has a certain value, would then come from a state-conditioned stimulus-response distribution. This would be a smaller region of stimulus-response space than the region from which all samples come (Fig. 1c). Therefore, if downstream neurons could respond in a state-dependent manner (contrary to common neural coding assumptions \cite{moreno2014information, stringer2019high}), state-dependent encoding would increase the capacity for discriminable stimulus encoding.

Characterisation of response distribution structure is also complicated by neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}, and in barrel cortex to repetitive whisker stimulation for spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}. That is, the response distributions are \textit{non-stationary} and change as a function of trial index for some neurons. Adaptation could therefore lead to correlative structure in single stimulus response distributions. Adaptation must therefore be controlled for when searching for correlative structure indicative of a possible underlying state.

The topographic representation of each single whisker on the snout of a rodent in the somatosensory cortex, the so-called barrel cortex, represents an excellent model to study neuronal processing of defined sensory stimulation \cite{feldmeyer2013barrel}. The shortness of whisker deflections reduces ambiguity in what temporal aspect of a stimulus neural activity represents, offering controlled study of the neural representation of atomic entities. The majority of barrel cortex neurons respond 0 or 1 times to single whisker deflections \cite{reyes2014laminar}, implicating single spike patterns as a candidate for atomic representation of information. Representation by single spike patterns is supported in other sensory modalities, such as vision, where a recent optogenetics study showed that early visual cortex patterns of single spikes are sufficient to enable visual discrimination \cite{resulaj2018first}.

First spike times with millisecond order precision have been observed relative to stimulus onset in sensory organs \cite{johansson2004first, uzzell2004precision, gollisch2008rapid}, thalamic areas \cite{reinagel2002precise, storchi2012comparison} and early sensory cortices \cite{reyes2014laminar}, as well as relative to oscillations \cite{havenith2011synchrony}. Relative differences \cite{izhikevich2006polychronization} observed experimentally \cite{johansson2004first, gollisch2008rapid, reyes2014laminar} have been shown to encode information when trial averaged \cite{gollisch2008rapid} and do not require reference to an onset signal.
Response distributions may form multiple clusters, implying stochasticity in the cluster from which representations come on a single trial (Fig. 1d; \cite{reyes2014laminar}).
% as neurons are sensitive to the spatial and temporal pattern of synaptic activations \cite{gasparini2006state, branco2010dendritic, branco2011synaptic}.

Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, first spike response distributions were first observed for single stimuli and pairs of units. That is, for repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked. The reader is strongly recommended to view these plots in the top left of Supplementary Vid. 1. Simple observation of these plots (ignoring lines, colours and ellipses at this stage) shows that clusters of first spike pair responses appear correlated, suggesting that first spike pairs are correlated for pairs of neurons on a subset of trials of a stimulus.

% That is, each axis corresponds to the first spike time of a different recorded unit. The scatter points therefore represent the first spike times of both units for the trials (of a single stimulus) on which both units spiked.

Moreover, the correlation angle of some of the correlated clusters appears to be different from $45^{\circ}$ (i.e. has a gradient different from 1).
A simple shift in the precise timing of stimulus onset or the onset of cortical activity, would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle). The observed correlations with `non-unitary gradient' may suggest that correlations are due to an internal process, for example, a shared cortical state underlying state-dependent processing. Such correlations with `non-unitary gradient' must be considered after checking and controlling for neural adaptive trends.

A novel algorithm extracted the observed clusters and found that they were correlated above chance levels. To control for adaptive trends, the cluster spike times of each neuron in the pair were tested for stationarity. The cluster spike times of both units were stationary for X\% of correlated clusters. X\% of these stationary correlated clusters had non-unitary (different from $45^{\circ}$) correlation angles, suggestive of an internal underlying state co-varying first spike times and thus \textit{warping} the distribution of (decodable) relative spike time differences from trial-to-trial (Fig. 1d).

For the remaining clusters, with at least one unit with non-stationary cluster spike times, time series analysis was applied to remove the neural adaptive trends of the non-stationary unit/units. X\% of the detrended clusters were correlated with non-unitary gradient, suggestive of an underlying state.

Application of state analysis to the stationary and detrended clusters provided an estimate of possible state-dependent response distributions. This gave a prediction of the resolution of possible state-dependent encoding, which would surpass previously proposed coding forms and controls for neural adaptive trends. It also provides a precise understanding of stimulus responses and correlative variability, which could be interpreted classically (i.e. independently of state [Vicente Refs]).

The precise characterisation of correlative structure in this paper is a necessary component step towards either rejecting state-dependent coding, or confirming its precise nature. It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons. Testing the state-dependent decoding performance of the response properties uncovered in this paper would require explicit knowledge (or an estimate) of the underlying trial-to-trial state so is left for future studies.
Subthreshold membrane potentials are a candidate for such state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}.

The characterised response distribution structure is referred to as `stationary and non-stationary clustered time-warped spike timing'. This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, non-stationarity and different warping functions for different neurons \cite{williams2020discovering}. One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons need not respond identically to warps of the same pattern. Rather, neurons should process relative spike patterns in a state-dependent manner.

The results suggest that consideration of clustering, state-dependence and non-stationarity, together, are necessary to explain a large proportion of neural variability, and could explain the failure to find a fast and efficient ubiquitous spike code that utilises the high sensitivity of neural architecture and functional dynamics. In this study, such consideration gave a novel prediction of the resolution of neural stimulus representation.

>>>>

\title{evidence of adaption-independent time-warped spike timing noise correlations in cortical networks}

@@

\section{Introduction}
The neural coding problem asks how information is encoded through patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}. Whilst precise temporal patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of spike time variability remains unexplained. Such variability leads to the common assumption of population codes based on spike counts \cite{reyes2015high, lin2015nature} or trial-averaged firing rates \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}.
% \cite{gerstner1997neural}, which have capacity beyond the original spike count \cite{}. 
% Continuous firing rates (averaged over time, trial or population ) 
% These population codes assume that trial-to-trial variability is either independent or correlated across the population \cite{averbeck2006neural, lin2015nature}. 
Trial-to-trial spike count variability is in fact correlated across the population in early sensory cortices \cite{reyes2015high, lin2015nature} and was shown to be modulated by a two-dimensional continuous state \cite{lin2015nature}. It is possible that spike time variability is modulated across a population by the same or similar low-dimensional underlying state. This study aims to characterise the nature of trial-to-trial spike time correlations as a step towards confirming or rejecting the modulation of spike times by an underlying low-dimensional latent state. This offers a novel prediction of the encoding capacity of state-dependent spike time encoding. Characterising spike timing codes is important to understanding a variety of dynamic, structural, homeostatic and cognitive processes, as well as properties of action potential patterns directly involved in behaviour.

% Continuous firing rates (averaged over time, trial or population \cite{gerstner1997neural}) have capacity beyond the single trial discrete spike count from which they are calculated, leaving a full understanding of neural coding and variability undetermined.

% , under the premise that variability cancels out across a population \cite{abbott1999effect, moreno2014information, stringer2019high} or that variability is shared across the population and modulated from trial-to-trial by some underlying low-dimensional latent state \cite{lin2015nature}.

% Characterising responses may also uncover response properties involved in behaviour.

Responses following sensory stimulation can be analysed to explore such principles of neural coding. 
% Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus. 
Particularly, a multi-neuron response can be expressed as a point in multi-dimensional space, where each dimension represents the value of a spike train metric. Responses from multiple presentations of a stimulus form a sample stimulus-response distribution in this space, which can be used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig. 1a).
% Estimating the underlying response distribution (including correlative structure) for single stimuli is complicated by neural adaptation and heterogeneity of sample response distribution structure.
% which approximates an underlying stimulus-response distribution.

@@

% Information theoretic approaches aim to understand 
If two stimuli have a behaviourally relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus. This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig. 1a). Neural coding is classically interested in determining the dimensions in which response distributions do not overlap \cite{averbeck2006neural}.

Under such approaches, `noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimulus \cite{averbeck2006neural}. `Signal correlations' concern tuning similarity between neurons over multiple stimuli (see \cite{averbeck2006neural}). As discussed, previous studies have demonstrated that spike count noise correlations are modulated across a population  by low-dimensional continuous latent states \cite{lin2015nature}. 
% For example, a 2-dimensional continuous latent state modulates responses in V1 \cite{lin2015nature}.

Neural coding and noise correlations are classically considered with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}. It is important to consider, that if downstream neurons are also modulated by the same population wide continuous latent state, then downstream `decoder' neurons could respond in a state-dependent manner.
When the continuous latent state has a certain value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig. 1c). State-dependent decoding would therefore offer an improved encoding capacity, compared to the classical fixed-state decoder interpretation.

% Correlative structure between the responses of different neurons in single stimulus response distributions indicates the presence of trial-to-trial noise shared between the responses of multiple neurons. These are classically referred to as `noise correlations' \cite{averbeck2006neural}, rather than

% reid2012functional
% It is important, however, to consider that trial-to-trial noise correlations between spatially distanced neurons could be caused by an underlying \textit{continuous} single or low-dimensional latent state that varies from trial-to-trial.  Such state-dependent coding would depend on the ability of downstream neurons to respond in a state-dependent manner. This could be possible if downstream neurons were also modulated by the same underlying continuous latent states. 
% Such state-dependent coding differs from the classical neural coding assumption of fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}).

% This paper aims to characterise the precise nature of spike timing noise correlations for single stimuli to uncover the potential encoding capacity of a state-dependent spike time code. 
% The stimulus-dependency of noise correlations is left to future studies.  
Characterising noise correlations within stimulus-response distributions is complicated by the heterogeneity of sample response distribution structure and neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}. 
% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
Processes of adaptation, which we define as different to latent states underlying trial-to-trial noise correlations, could also lead to correlative structure in the stimulus-response distribution (Fig. X). Adaptation must therefore be controlled for when characterising adaption-independent noise correlations.

The topographic representation of each single whisker on the snout of a rodent in the somatosensory cortex, the so-called barrel cortex, represents an excellent model to study noise correlations for controlled sensory stimulation \cite{feldmeyer2013barrel}. 
% The shortness of whisker deflections reduces ambiguity in what temporal aspect of a stimulus neural activity represents. 
The majority of barrel cortex neurons respond 0 or 1 times to single whisker deflections \cite{reyes2014laminar}, implicating single spike patterns as a candidate for encoding. Single spike pattern encoding is supported in other sensory modalities, such as vision, where patterns of single spikes are sufficient to enable visual discrimination \cite{resulaj2018first}.

The nature of first spike patterns has been explored in previous research. First spike times have been observed with millisecond order precision relative to stimulus onset across species and modalities in sensory organs \cite{johansson2004first, uzzell2004precision, gollisch2008rapid}, thalamic areas \cite{reinagel2002precise, storchi2012comparison} and early sensory cortices \cite{reyes2014laminar}, as well as relative to oscillations \cite{havenith2011synchrony}. First spike times in the rat barrel cortex have been shown to improve stimulus encoding by 44\% over spike count \cite{panzeri2001role}. Precise relative differences between the spike times of multiple neurons \cite{izhikevich2006polychronization} do not require reference to an onset signal and have been observed experimentally \cite{johansson2004first, gollisch2008rapid, reyes2014laminar} and shown to encode stimulus information when averaged over trials \cite{gollisch2008rapid}.

Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of spike time noise correlations for first-spike patterns, towards estimating the potential encoding capacity of a state-dependent spike time encoding and determining the dimensionality of such an underlying latent state. The classical `stimulus-dependency' of noise correlations is left to future studies. Previous studies analysing the same dataset found precise multi-neuron intralaminar spike patterns \cite{reyes2014laminar}, spike count noise correlations and could decode stimulus properties (stimulation frequency and locaction) from spike counts and spike patterns \cite{reyes2015high}. The new analysis uses time series analysis to control for neural adaption over all stimulus trials. Adaption occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}, and was confirmed to occur for 7 out of 8 neuron groups in this study.

% as neurons are sensitive to the spatial and temporal pattern of synaptic activations \cite{gasparini2006state, branco2010dendritic, branco2011synaptic}.

For repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked. The reader is encouraged to view these plots in the top left of Supplementary Vid. 1. A novel algorithm extracted isolated clusters from these single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the data set \cite{reyes2014laminar}. The novel algorithm, however, also showed that first spike pairs within these isolated clusters were positively correlated above chance-levels.

Moreover, the correlation angle of many of the correlated clusters was different from $45^{\circ}$. This cannot be explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle). Instead the `non-unitary' correlation gradients (different from $45^{\circ}$) may be indicative that correlations are due to an internal shared latent state. Clusters remained correlated with non-unitary gradient after controlling for neural adaption using time series analysis, confirming the correlated clusters as evidence of precise non-adaptive noise correlations. A large proportion of these noise correlations would have been missed without clustering.

Importantly, the observation of adaption-independent correlated clusters with non-unitary gradient shows that the distribution of decodable first spike time relative differences for pairs of neurons are `warped' from trial-to-trial by some `non-adaptive' continuous latent state. This phenomenon is referred to as `adaption-independent clustered time-warped spike timing'. This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, adaption and different warping functions for different neurons \cite{williams2020discovering}. One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons need not respond identically to warps of the same pattern. Rather, neurons should process relative spike patterns in a state-dependent manner to utilise the improved state-dependent coding capacity.

Adaption-controlled non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers. This offers suggestive evidence of a shared low-dimensional continuous latent state affecting the spike times of spatially distanced neurons, which in turn could also modulate downstream neurons and allow them to respond in a state-dependent manner. Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaption-controlled correlated cluster. This gave a prediction of the resolution of possible state-dependent encoding, which surpasses previously proposed coding forms.

The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed. Estimating whether cluster correlations observed for different pairs of neurons is modulated by a single or low-dimensional shared state was made difficult by variability between clusters in the trials from which cluster samples came [See notes below]. The work of this paper in extracting and characterising adaption-independent correlated clusters provides a necessary component step towards confirming or rejecting state-dependent spike-time coding, and is also important under the classical fixed-state decoder interpretation. The results show that consideration of clustering, correlative structure and adaption, together, are necessary to explain a large proportion of neural variability.

% It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons. 
Testing the state-dependent decoding performance of the clusters is left for future studies, as it would require explicit knowledge of the underlying shared state or an estimate from the spike times of other neurons (which is difficult for the current extracted cluster counts per stimulus in single experimental sessions).
Subthreshold membrane potentials would be a candidate for such state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}. The LFP may offer a useful proxy for the subthreshold membrane potential.

% and could explain the failure to find a fast and efficient ubiquitous spike code that utilises the high sensitivity of neural architecture and functional dynamics. In this study, such consideration gave a novel prediction of the resolution of neural stimulus representation.

\subsection{Notes}
I have begun developing several techniques to address this but I think it may prove difficult to show anything strongly conclusive either way. I can give more details of this once we are agreed on the overall objectives. We may be able to address this issue, but it may turnout to require future experiments designed to get a higher cluster count per stimulus and experimental session. A bit of help attacking this problem and working on factor analysis in 3+ dimensions would also be helpful.
